---
title: "NICEI - Northern Ireland Composite Economic Index"
output: 
  github_document:
    toc: true
    toc_depth: 4
date: "2023-04-19"
editor_options: 
  markdown: 
    wrap: 72
always_allow_html: true

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = FALSE, warning = FALSE, message = FALSE)
```

This document will outline the steps taken, including the code, to
achieve the plots in the "Growth Rates" column of the NICEI Q4 2022
factsheet which can be found
[here](https://www.nisra.gov.uk/system/files/statistics/NICEI%20Q4%202022%20factsheet.PDF).

# Growth Rates

## 1. NICEI Trend to Q4 2022

#### Step 1: Installing relevant libraries

```{r }
library(readxl)
library(janitor)
library(ggplot2)
library(dplyr)
library(reshape2)
library(tidyverse)

```

#### Step 2: Data Import and Preparation

##### Loading data

Import from Excel spreadsheet in local directory

```{r Data Import}
df <- read_excel("NICEI-Tables-Q4-2022.xlsx", sheet = "Table 1", skip = 1)
```

##### Data Preparation

###### Data Selection

Dropping last 3 columns that are unused and merging the Year and Quarter
column

```{r Data Cleaning}
df1 <- df %>% 
  select(c(0:5)) %>%
  mutate(date = paste(Year, Quarter, sep="\nQ"), .before = 1, .keep = "unused")

```

###### Data Restructuring

Now for restructuring. This code takes the data from the 4 columns and
restructures it to 3 columns which contain the date, the variable
(previous column headers i.e the variables of interest) and their
respective values. This is useful for colouring plots based on category
as is needed in this case.

```{r}
df2<-melt(df1, id.vars="date") 
```

#### Step 3. Plotting

Now to plot the data to find out what it looks like and what more needs
done

##### Initial plot

```{r initial plot}
#plot of data
ggnicei<-ggplot() + 
  geom_line(data = df2, 
            aes(x=date, y=value, 
                group = variable, 
                color = variable)
            )
ggnicei
```

The plot lines look correct but work needs done to make the legend
legible and prettify the plot so it resembles the original plot
published. To alter the X-axis values so they're readable, a function
will be created to return a function as the breaks argument; this, in
turn, will return the appropriate x-axis spacing i.e the nth value
specified when the function is called.

##### Plot formatting/aesthetics

```{r function}

#Create function 
every_nth = function(n) {
  return(
    function(x){
      x[c(TRUE, rep(FALSE, n - 1))]
      }
    )
}

```

The plot can now be redone implementing some formatting such as:

-   scale_colour_manual: this formats the plot in relation to colour
-   geom_hline: this inserts a horizontal line
-   annotate: this permits annotation on the plot
-   theme: this has many possible components that all will affect the
    aesthetic of the plot

```{r final plot, fig.height=7, fig.width=10}

final_plot<-ggnicei + 
  scale_color_manual(values= c("black", "blue", "#cccc00"),             # selecting colours for the lines
                     labels=c("NICEI",                                  # creating the line labels for the legend
                              "Private Sector Component Index",         
                              "Public Sector Component Index")) +
  geom_hline(yintercept=100, linetype = 'dashed') +                    # generating the dashed horizontal line (baseline)
  annotate("text",                                                     # Forming the text for the baseline
           x="2014\nQ1", 
           y= 99, 
           label = "Baseline 2019 = 100", 
           size =3, 
           fontface = "bold") + 
  theme(line = element_blank(),                                 # plot theme
        panel.grid.major = element_blank(),                     # no major gridlines
        panel.grid.minor = element_blank(),                     # no minor gridlines
        panel.background = element_blank(),                     # blank background
        plot.title = element_text(colour = "chartreuse3",       # title formatting: colour and bold
                                  face = "bold"),
        axis.text = element_text(face = "bold"),                # formatting axis text: bold
        axis.line = element_line(colour = "NA"),                # no axis lines
        legend.title=element_blank(),                           # remove legend title 
        legend.key=element_blank(),                             # remove background colour from legend key
        legend.position = c(0.15,0.15),                         # legend position
        plot.margin = unit(c(2,2,2,2), "cm")) +                 # adjusting plot padding for exported image
  scale_x_discrete(breaks = every_nth(n = 8)) +                 # implementing the function for axis labels every 8th (Q1 every other year)   
  scale_y_continuous(breaks = seq(75, 110, by=5),               # formatting Y-axis: 75-115 in increments of 5
                     limits=c(75,110))+
  annotate(geom = "point",      # inserting red point at Y-value 105.7
           x = "2022\nQ4",      
           y = 105.7, 
           colour = "red", 
           size = 2) +
  annotate(geom = "label",      # inserting blue label with white font stating 105.7
           x = "2021\nQ4", 
           y = 108, 
           label = "105.7", 
           hjust = "left", 
           fill="blue", 
           color = "white") +
  annotate("segment",           # connecting line between point and label
           x = "2022\nQ1", 
           xend = "2022\nQ4", 
           y = 107.8, yend = 105.7,
           linewidth = 0.2) +
  labs(x = "",                                  # no X-axis title
       y="",                                    # no Y-axis title
       title = "NICEI Trend to Q4 2022")        # plot title

final_plot

```

This plot seems much more comparable to the original in the publication;
however, the plot in the original pdf seems to have a thicker NICEI
line. If this is desired, all that is required is to overlay the
final_plot with the same plot line but to the required thickness.

```{r thicken plot line, fig.height=7, fig.width=10}
df4<-filter(df2, variable == "NICEI")   # extract the values associated with the required plot line 
final_plot2<- final_plot +              # add these extracted values to the previous plot, formatting appropriately
  geom_line(data=df4, 
            aes(x=date, 
                y=value, 
                group = 1), 
            linewidth = 1)
final_plot2
```

#### Step 4. Export plot image

Instead of exporting the plot as an image manually within R Studio, a
scripted export of plot image can be done; however, it must be noted
that there is a decrease in text readability using this method and so
will need increased. An example of the manually exported plot is in the
local directory named, "NICEI_r\_plot.png". Scripted export can be done
as follows:

```{r results = "hide"}

png("NICEI-plot.png", 
    width = 950, 
    height = 750)
final_plot2
dev.off()
```

## 2. NICEI Trend to Q4 2022

#### Step 1: Installing relevant libraries

```{r libaries}
library(readxl)
library(xts)
library(formattable)
library(flextable)
library(tidyverse)
library(gridExtra)
library(grid)
library(officer)
```

#### Step 2: Data Import, Cleaning and Preparation

##### Loading data

Read in data (excel format) from local directory

```{r}
df <- read_excel("NICEI-Tables-Q4-2022.xlsx", sheet = "Table 1", skip = 1)
formattable(head(df))
```

```{r}
#calculate the data required for the Growth table from the raw data (fractional differences)
Q_Q <- (df[nrow(df),] - df[nrow(df)-1,])/df[nrow(df)-1,]    # Calculate Q/Q row
Y_Y <- (df[nrow(df),] - df[nrow(df)-4,])/df[nrow(df)-4,]    # Calculate Y/Y row
Tri <- (df[nrow(df),] - df[nrow(df)-12,])/df[nrow(df)-12,]  # Calculate Triennial row
```

```{r}
# Create new column with year and Quarter combined, separated with a space, and delete the two individual columns 
df1<- df %>% mutate(date = paste(Year, Quarter, sep="\nQ"), .before = 1, .keep = "unused")

# Convert from data frame to a time series object to enable use of moving/rolling average calculations
TS <- df1 %>% ts(frequency = 4, start = c(2006,1), end = c(2022, 4))    # frequency = 4 means data consists of quarterly samples.

# Calculate the 4Q rolling average of dataset
r4Q <-TS %>%                   # all data (window() would be used if calculations were to be on data subsets)
  as.xts() %>%                 # coerces the timeseries object to an eXtensible format required for time-based analysis
  apply.yearly(mean) %>%       # apply mean function to each distinct year of data
  as.data.frame()              # pipe data to data frame for additional work

# calculation of fractional difference for rolling 4Q annual average
r4q<-(r4Q[nrow(r4Q),] - r4Q[nrow(r4Q)-1,])/r4Q[nrow(r4Q)-1,]

# create the base table of Growth rates
dt<-rbind(Q_Q[,-c(1:2)],
          Y_Y[,-c(1:2)], 
          Tri[,-c(1:2)], 
          r4q[,-1]) %>%   
  round(digits = 3)

# create the column of names
growth <- data.frame("Growth Rates" = c("Q/Q",
                                        "Y/Y", 
                                        "Triennial \n(3Yr) Change", 
                                        "Rolling 4Q \nAnnual Ave")
                     )

DT<-cbind(growth, dt)   # combine base table with names
colnames(DT)<-c("growth", "NICEI", "private", "public", "services", "production", "construction")  #convert column names to single value names (needed for the compose later)
DT


pDT<- DT %>% mutate(across(-c(1),percent))    #convert values to percentage
dft<-flextable(pDT) %>% theme_box()   # pretty print of table using flextable (easier to customise than formattable)

#prettify table
dft<-bold(dft, part = "header")                   # bold header
dft<-bg(dft, bg = "olivedrab3", part = "header")  # fill the header with colour
dfxt <- dft                                       # rename for next step of adding arrows  
dfxt

```

Complex process of adding coloured arrows corresponding to values. Each
column must be addressed first adding the green arrows, as this is the
predominant colour Then each column with red arrows must be edited to
reflect this. This is a manual process because there isn't a consistent
rule upon which to allocate the arrows. Each value corresponds to it's
previous value in a different year/quarter dependent on the category
it's describing. There are R emoticon/graphics packages available but it
was easier to find .png files and save them to the local directory and
upload

```{r}
dfxt2 <- flextable::compose(dfxt,   #data
                 j = 2,   #2nd column
                 value = as_paragraph(     # allows concatentation of text chunks and inmages etc
                   as_image(src = "arrow_up.png", width = .15, height = .15),   # arrow first as per the publication being replicated
                   " ",    # space follows the arrow
                   as_chunk(NICEI, props = fp_text(color = "black")    # The text to be added, which is the value in the column NICEI. 
                            )                                           # This line is why the column names needed changed earlier (line 49)
                   ),
                 part = "body")   # noting that the compose function is applied to the table body
dfxt2    #table check
dfxt2 <- flextable::compose(dfxt2,
                 j = 3,
                 value = as_paragraph(
                   as_image(src = "arrow_up.png", width = .15, height = .15), " ",
                   as_chunk(private, props = fp_text(color = "black"))
                 ),
                 part = "body")
dfxt2
dfxt2 <- flextable::compose(dfxt2,
                 j = 4,
                 value = as_paragraph(
                   as_image(src = "arrow_up.png", width = .15, height = .15), " ",
                   as_chunk(public, props = fp_text(color = "black"))
                 ),
                 part = "body")
dfxt2
dfxt2 <- flextable::compose(dfxt2,
                 j = 5,
                 value = as_paragraph(
                   as_image(src = "arrow_up.png", width = .15, height = .15), " ",
                   as_chunk(services, props = fp_text(color = "black"))
                 ),
                 part = "body")
dfxt2
dfxt2 <- flextable::compose(dfxt2,
                 j = 6,
                 value = as_paragraph(
                   as_image(src = "arrow_up.png", width = .15, height = .15), " ",
                   as_chunk(production, props = fp_text(color = "black"))
                 ),
                 part = "body")
dfxt2
dfxt2 <- flextable::compose(dfxt2,
                 j = 7,
                 value = as_paragraph(
                   as_image(src = "arrow_up.png", width = .15, height = .15), " ",
                   as_chunk(construction, props = fp_text(color = "black"))
                 ),
                 part = "body")
dfxt2
dfxt2<-  flextable::compose(dfxt2,
                 j = 6, i = ~production < 0.011, 
                 value = as_paragraph(
                   as_image(src = "arrow_down.png", width = .15, height = .15), " ",
                   as_chunk(production, props = fp_text(color = "black"))
                 ),
                 part = "body")
dfxt2
dfxt2 <- flextable::compose(dfxt2,
                 j = 7, i = ~construction < 0.04,
                 value = as_paragraph(
                   as_image(src = "arrow_down.png", width = .15, height = .15), " ",
                   as_chunk(construction, props = fp_text(color = "black"))
                 ),
                 part = "body")

dfxt2   #this should be the completed body 

```

```{r}
#Need to change the header names back to the proper format
dfxt3 <- set_header_labels(dfxt2,
                           values = list(
                             growth = "Growth Rates",
                             NICEI = "NICEI",
                             private = "Private Sector",
                             public = "Public Sector",
                             services = "Services",
                             production = "Production",
                             construction = "Construction"
                           )
)

dfxt3
```

```{r}
#prettifying the table via formatting to reflect the publication be replicated
dfxt4 <- width(dfxt3, j = 1, width = 1.3086) %>%    # Adjusting the width of column 1 
  align(align = "center", part = "all") %>%         # Center aligning all of table contents
  align(j=1, align = "left", part = "body")         # Left aligning column 1

dfxt4
growth_table <-dfxt4    # Final table
```
